---
title: "PS270 Final Project"
author: "Ryan Konet"
description: "The Effect of Domestic Migration on Voting Patterns"
output:
  distill::distill_article:
    self_contained: false
---



## Introduction
It is a common political complaint among Americans in more conservative areas that when people move in from out of state they "take their politics with them." The three states with the largest exodus of population are solidly blue states with large blue metropolitan areas: California, Illinois, and New York. But is this trope actually true or is it borne out of some sort of large scale "survivor-ship bias"? In this research, each county in 48 states (excluding the District of Columbia, Alaska, and Connecticut for statistical reasons) will be used to compare their changes in voting patterns from the 2020 and 2024 Presidential Elections and see if the proportionate net migration between those two elections had any effect on voting patterns. This project will seek to answer whether the flow of domestic migration into an area has a major effect on electoral behavior or, at minimum, if they coincide. Put more concisely, as the net domestic migration rate becomes proportionally greater in any given county, the share of votes going toward the GOP decreases on average. This research question and hypothesis is important to investigate because, if true, it allows political scientists and campaigns to better predict the changes a region could see to its partisan politics between election cycles so that they can adapt their studies accordingly. It is important to also note that this is an observational cross-sectional study. 

## Data Section, Sources and Variables
The data used in this experiment is a combination of three sets: 2020 Presidential Election Results by County, 2024 Presidential Election Results by County, and Annual Resident Population Estimates, Estimated Components of Resident Population Change, and Rates of the Components of Resident Population Change for States and Counties: April 1, 2020 to July 1, 2024 (CO-EST2024-ALLDATA). The 2020 Presidential Election Results were taken from a github user (https://github.com/tonmcg/US_County_Level_Election_Results_08-24/blob/master/2020_US_County_Level_Presidential_Results.csv). The same can be said for the 2024 Presidential Election Results (https://github.com/tonmcg/US_County_Level_Election_Results_08-24/blob/master/2024_US_County_Level_Presidential_Results.csv).The 2020 to 2024 Population Changes were taken from the United States Census Bureau. (from "Datasets" at the bottom of https://www.census.gov/data/tables/time-series/demo/popest/2020s-counties-total.html).

These datasets were all merged into one dataset called `focused_per_gop`. Below are the variables that were taken or created out of these datasets. There were issues with identification numbers in Connecticut, Alaska, and Washington DC, so they are excluded from this research.

| Name           | Description                                               |
|:---------------|:----------------------------------------------------------|
| `state_name`   | name of the state the county belongs to      |
| `county_fips`       | county identification number                         |
| `county_name`   | name of the county       |
| `per_gop_2020` | The proportion of votes cast for a GOP president in the 2020 Election    |
| `per_gop_2024` |  The proportion of votes cast for a GOP president in the 2024 Election     |
| `per_gop_change` |  The change in the percent of GOP votes from 2020 to 2024   |
| `per_gop_change_adjusted` |  per_gop_change adjusted for national difference in election results   |
| `dom_mig_total`   | The net domestic migration from 2020 to 2024 |
| `dom_mig_per`   | The proportion of net domestic migration by 2020 population |
| `population_2020` | population in 2020 |


## Data Section, Variables and Their Relationships
The key variables that are used in the relationship are `dom_mig_per` and both `per_gop_change` and `per_gop_change_adjusted` variables. The independent variable being the former and dependent the two latter. The `dom_mig_per` adds the net domestic migration from 2020 to 2024 together and then divides by the `population_2020` variable for each county. The two dependent variables are the exact same except `per_gop_change_adjusted` has every datum subtracted by 0.0295, which is the difference between the national GOP vote shares in each election. The difference between Democrat vote shares is 0.0299, so the third party shift is negligible enough to ignore in the adjustment. The adjusted dependent variable will be the preferred method of outcome; however, the non-adjusted version will usually be presented alongside it. The relationship measures how the proportion of net domestic migration affects the proportion of change in GOP vote share in the majority of the voting counties. Success of this experiment will be measured by a statistically significant negative relationship between these two variables. 

Below are two histograms that show the distribution of the outcome variable(s) (dependent) `per_gop_change` and `per_gop_change_adjusted` respectively.The major difference between the two being that the distribution of the adjusted is, naturally, much more centered on a zero percent increase. Keep in mind that in both graphs there exists some natural skewing because in a regular count a county with 8,000 people is counted the same as a county with 1,000,000. These graphs are just meant to show that there exist changes within the years whether accounting for the adjustment or not.

Below those two histograms is a table that lists the 48 states and sorts them by net domestic migration. States are used here because just using counties would be far too long and it can illustrate some patterns that will be revisited in the results section going by county, although right away it appears the correlation is not bared out on state level by seeing 3 of the top 5 states by proportion of net domestic migration actually gaining GOP "voteshare" even when adjusted. 

The data will be interpreted as a cross-sectional study due to the lack of repetition, treatment, randomness, et cetera.

```{r,echo=FALSE}
library(tidyverse)
  # 2020 Presidential Election
pres_2020 <- read_csv("pres_data/2020_US_County_Level_Presidential_Results.csv")
pres_2020 <- pres_2020 |> filter(state_name != "Alaska", state_name != "Connecticut", 
                                 state_name != "District of Columbia")
pres_2020 <- pres_2020 |> select(-votes_dem, -per_dem, -diff, -per_point_diff) |> 
  rename(votes_gop_2020 = votes_gop, per_gop_2020 = per_gop, total_votes_2020 = total_votes)
  # 2024 Presidential Election
pres_2024 <- read_csv("pres_data/2024_US_County_Level_Presidential_Results.csv")
pres_2024 <- pres_2024 |> filter(state_name != "Alaska", state_name != "Connecticut", 
                                 state_name != "District of Columbia")
pres_2024 <- pres_2024 |> select(-votes_dem, -per_dem, -diff, -per_point_diff) |> 
  rename(votes_gop_2024 = votes_gop, per_gop_2024 = per_gop, total_votes_2024 = total_votes)

  # 2020-2024 Population Estimates
pop_2024 <- read_csv("pop_data/co-est2024-alldata.csv")
pop_2024 <- pop_2024 |> filter(STNAME != "Alaska", STNAME != "Connecticut", STNAME != "District of Columbia")
pop_2024 <- pop_2024 |> mutate(county_fips = paste0(STATE, COUNTY)) |> 
  mutate(dom_mig_total = (DOMESTICMIG2020 + DOMESTICMIG2021 + DOMESTICMIG2022 + DOMESTICMIG2023 + DOMESTICMIG2024)) |> 
  mutate (dom_mig_per = dom_mig_total / ESTIMATESBASE2020) |> mutate(population_2020 = ESTIMATESBASE2020)
```
```{r, echo=FALSE}
  # 2020-2024 Population Estimates
pop_2024_important <- pop_2024 |> select(county_fips, dom_mig_total, dom_mig_per, population_2020)
pop_2024_50_states <- pop_2024 |> filter(STNAME == CTYNAME) |> 
  arrange(desc(dom_mig_per)) |>
  select(STNAME, dom_mig_per, population_2020) 

  # Data Merging
percent_gop <- left_join(pres_2020, pres_2024) |> mutate(per_gop_change = per_gop_2024 - per_gop_2020)
percent_gop <- left_join(percent_gop, pop_2024_important, by = "county_fips")
percent_gop <- percent_gop |> mutate(per_gop_change_adjusted = per_gop_change - 0.0295)
focused_per_gop <- percent_gop |> select(state_name, county_fips, county_name, per_gop_2020, per_gop_2024, per_gop_change, per_gop_change_adjusted, dom_mig_total, dom_mig_per, population_2020)

  # State Data
percent_gop_50_states <- percent_gop |> group_by(state_name) |> 
  summarize(dom_mig_total = sum(dom_mig_total), population_2020 = sum(population_2020), gop_2020 = sum(votes_gop_2020),
            gop_2024 = sum(votes_gop_2024), total_2020 = sum(total_votes_2020), total_2024 = sum(total_votes_2024))
percent_gop_50_states <- percent_gop_50_states |> 
  mutate(domestic = (dom_mig_total / population_2020), gop_2024_per = (gop_2024 / total_2024), 
         gop_2020_per = (gop_2020 / total_2020)) |> 
  mutate(gop_change = (gop_2024_per - gop_2020_per))
percent_gop_50_states <- percent_gop_50_states |> 
  mutate(dom_mig_per = dom_mig_total/population_2020, gop_change_adj = gop_change - 0.0295) |> 
  select(state_name, dom_mig_per, gop_change_adj, gop_change, population_2020) |> arrange(desc(dom_mig_per))
```

```{r, echo=FALSE}
plot_dep_var <- focused_per_gop |> ggplot(mapping = aes(x = (per_gop_change * 100))) + 
  geom_histogram(binwidth = 0.5) +
  labs(title = "Change in GOP Vote Share by County from 2020 to 2024", 
       x = "Change in GOP Voteshare per County, 2020-2024 (percentage points)", 
       y = "Number of Counties", 
       caption = "Sources: 2024 PEP Census Data, 2020 and 2024 US House of Representatives Election Statistics") + 
  theme(axis.title.x = element_text(size = 8))
plot_dep_var

plot_dep_var_change <- focused_per_gop |> ggplot(mapping = aes(x = (per_gop_change_adjusted * 100))) + 
  geom_histogram(binwidth = 0.5) + 
  labs(title = "Change in GOP Vote Share (adjusted) by County from 2020 to 2024", 
       x = "Change in GOP Voteshare per County (adjusted), 2020-2024 (percentage points)", 
       y = "Number of Counties", 
       caption = "Sources: 2024 PEP Census Data, 2020 and 2024 US House of Representatives Election Statistics") + 
  theme(axis.title.x = element_text(size = 8))
plot_dep_var_change
```


```{r, echo=FALSE}
 # Table for all 50 states
knitr::kable(percent_gop_50_states, col.names = c("State", "Net Domestic Migration Proportion", "GOP Voteshare Change Adjusted", "GOP Voteshare Change", "2020 Population"))
```


## Results Section: 
This first graph simply plots each county in every state except for Alaska and Connecticut based on their net proportional domestic migration between the 2020 and 2024 elections and the change in the percent of votes that went to the GOP from the 2020 election to the 2024 election. The dashed line represents no change in vote share and the blue line represents the line of best fit. This graph has a clear negative correlation, meaning that as net proportional domestic migration increases the change in percent votes for the GOP in that county is, on average, likely to decrease. There are some flaws in this graph that hinder a perfect representation. The first flaw is that the national vote share for the GOP has not remained constant between 2020 and 2024; in fact, it increased by 2.95 percentage points.

```{r, echo=FALSE}
plot_basic <- ggplot(focused_per_gop, mapping = aes(x=(dom_mig_per * 100), y=(per_gop_change * 100))) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE, linewidth = 2.0) + 
  geom_abline(intercept = 0.00, slope = 0, color = "grey", linetype = "dashed") + 
  labs(title = "The Effect of Domestic Migration on the 2024 Presidential Election", 
       x = "Net Proportional Domestic Migration per County, 2020-2024 (percentage of 2020 population)", 
       y = "Change in GOP Voteshare per County, 2020-2024 (percentage points)", 
       caption = "Sources: 2024 PEP Census Data, 2020 and 2024 US House of Representatives Election Statistics") + 
  theme(axis.title.x = element_text(size = 8), axis.title.y = element_text(size = 7))
plot_basic
```

This graph adjusts the former graph by showing the change of the percent of GOP votes from the national average. In other words, this value is found by subtracting the GOP percent in 2020 from 2024. This value is 2.95 percentage points (the same calculation for Democrats was -2.99 percentage points, so the third party share's change was negligible). This table shows the exact same relationship as the previous graph, it is just adjusted so the vast swing of the national vote share doesn't misrepresent the changes in each county. There is still another issue: not all these counties have the same population.

```{r, echo=FALSE}
plot_basic_adjusted <- ggplot(focused_per_gop, mapping = aes(x=(dom_mig_per * 100), y=((per_gop_change_adjusted) * 100))) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE, linewidth = 2.0) + 
  geom_abline(intercept = 0.00, slope = 0, color = "grey", linetype = "dashed") + 
  labs(title = "The Effect of Domestic Migration on the 2024 Presidential Election", 
       x = "Net Proportional Domestic Migration per County, 2020-2024 (percentage of 2020 population)", 
       y = "Change in GOP Voteshare per County adjusted for national change, 2020-2024 (percentage points)", 
       caption = "Sources: 2024 PEP Census Data, 2020 and 2024 US House of Representatives Election Statistics") + 
  theme(axis.title.x = element_text(size = 8), axis.title.y = element_text(size = 6))

  plot_basic_adjusted
```

This third graph does the same thing as the second, except the line of best fit is calculated by weighting each data point (county) by its population. This means counties with a greater population will be considered as more consequential than one with a smaller population. Basically, if one county is 10,000 and the other 100,000, then the latter is weighted more. This graph shows that the magnitude of the negative relationship is actually greater when adjusted for population (more negative slope for the line of best fit), meaning this trend is likely more common among more urban counties. 

```{r,echo=FALSE}
plot_basic_adjusted_no_points <- ggplot(focused_per_gop, mapping = aes(x=(dom_mig_per * 100), y=((per_gop_change_adjusted)*100))) + 
  geom_point() + 
  geom_smooth(aes(weight = population_2020), method = "lm", se = FALSE, linewidth = 2.0) + 
  geom_abline(intercept = 0.00, slope = 0, color = "grey", linetype = "dashed") +
  labs(title = "The Effect of Domestic Migration on the 2024 Presidential Election weighted by 2020 Population", 
       x = "Net Proportional Domestic Migration per County, 2020-2024 (percentage of 2020 population)", 
       y = "Change in GOP Voteshare per County adjusted for National Change, 2020-2024 (percentage points)", 
       caption = "Sources: 2024 PEP Census Data, 2020 and 2024 US House of Representatives Election Statistics")  + 
  theme(axis.title.x = element_text(size = 8), axis.title.y = element_text(size = 6), plot.title = element_text(size = 11))
  
plot_basic_adjusted_no_points
```

These three graphs each move closer to the conclusion that counties with meaningful negative net domestic migration tend to increase their percentage of GOP votes and counties with meaningful positive net domestic migration tend to decrease their percentage of GOP votes. These two phenomenon become more likely and greater in magnitude when the magnitude of the net domestic migration is greater. But is this bared out in a regression model?

## Results Section, Regression
```{r, echo=FALSE}
reg_gop <- focused_per_gop |> 
  mutate(dom_mig_per = dom_mig_per*100, per_gop_change = per_gop_change * 100, 
         per_gop_change_adjusted = per_gop_change_adjusted *100)

regression_1 <- lm(per_gop_change ~ dom_mig_per, data = reg_gop)
regression_2 <- lm(per_gop_change_adjusted ~ dom_mig_per, data = reg_gop)
regression_3 <- lm(per_gop_change_adjusted ~ dom_mig_per, data = reg_gop, weights = population_2020)


library(broom)

tab1 <- tidy(regression_1) |> mutate(plot = 1, model = "Unweighted (Unadjusted)")
tab2 <- tidy(regression_2) |> mutate(plot = 2, model = "Unweighted (Adjusted)")
tab3 <- tidy(regression_3) |> mutate(plot = 3, model = "Weighted (Adjusted)")

regression_table <- bind_rows(tab1, tab2, tab3)

regression_table <- regression_table |> mutate(term = ifelse(term == "(Intercept)", "0 Net Dom Migration", "Coefficient"))

regression_table <- regression_table |> select(plot, model, term, estimate, std.error, statistic, p.value)

knitr::kable(regression_table, col.names = c("Plot", "Weight and Adjust Status", "Term", "Estimate", "Standard Error", "Statistic", "P-Value"))


```
The regression model has lots of information that can explain the relevance of the three plots. The Plot corresponds with the order the plots appeared. The Term tells whether this is the intercept, meaning when `dom_mig_per` is zero and there is 0 net domestic migration, or the coefficient, representing how many percentage points the GOP voteshare are expected to change for every increase of one percentage point by `dom_mig_per` in each plot. The Estimate represents the values for the corresponding term. The Standard Error is the estimated standard deviation of each term. The statistic is calculated by the Estimate divided by the Standard Error, and the higher the magnitude of the value the more statistically significant the relationship. The p-value calculates how probable the data distribution would be under a null hypothesis (a hypothesis with no correlation), it is scaled 0 to 1. 

Each plot, respectively from first to third, has an intercept (when net domestic migration = 0) of 2.023 percentage points, -0.0927 percentage points, and -0.212 percentage points.The first plot estimates that the GOP voteshare will decrease 0.055 percentage points for every one percentage point increase in net domestic migration. The second plot does the same because the only adjustment was to the intercept. The third plot will decrease 0.149 percentage points for every percentage point in crease in net domestic migration. Now, looking at the calculated t-statistic, which since all are far above 2 (in magnitude) as seen in the table above can all be called statistically significant in their relationship. Furthermore, the p-value of 0 in all six terms is far below the 0.05 benchmark, which makes it so that we can confidently rule out the null hypothesis and continue to rule that all three relationships are statistically significant. 

These three plots show that there is a negative relationship that is statistically significant between the net domestic migration into a given county and the share of votes in a presidential election that go to a GOP candidate. Essentially, counties with very high net domestic migration are likely to experienced a substantial decrease in their share of votes going toward the GOP. Conversely, counties with high net domestic migration out of the county see a substantial increase. Unfortunately, these relationships cannot necessarily attribute causality. The limits of the study being cross-sectional, mean that their is no way to definitively rule out other confounders like demographics or immigration.

## Conclusion
It is clear that if a county has a high net increase in domestic migration that it will likely see a substantial decrease in its proportion of votes for the GOP. This means that, at least as it pertains to correlation, the conservative trope rings true. However, there are still many issues that could use clearing up when it comes to causality. It is possible that there are many variables that are causing either just the GOP vote share decrease or both the vote decrease and net domestic migration increase. Examples include economic factors like GDP per capita, demographic factors like race and immigration, policy changes in party platforms, and other unforeseen events like how one election was during a pandemic and the other was not. In order to try and get closer to proving causality all of these possible confounders should be put through the same study as the net domestic migration to see if any bare a similar relationship. It would also be helpful to look both at other presidential elections (2016 to 2020 or 2016 to 2024), and elections for senate or congress for similar year ranges. It would even help to survey the average domestic migrant to see if they are more liberal than the average person in their destination. With infinite time and money, there would be much to explore within this relationship; however, even without all these extra studies, it can still be seen that there is a negative correlation between net domestic migration and the GOP vote share change over the same time period. 
