---
title: "PS270 Final Project"
author: "Ryan Konet"
description: "The Effect of Domestic Migration on Voting Patterns"
output:
  distill::distill_article:
    self_contained: false
---



## Project Description
It is a common political complaint among conservative Americans that when people move in from out of state they "take their politics with them." The three states with the largest exodus of population are solidly blue states with large blue metropolitan areas: California, Illinois, and New York. But is this trope actually true or is it borne out of some sort of large scale "survivorship bias"? In this study, each county in 48 states (excluding the District of Columbia, Alaska, and Connecticut for statistical reasons) will be used to compare their changes in voting patterns from the 2020 and 2024 Presidential Elections. In order to get a holistic view, the data will be organized by state, population, et cetera.This project will seek to answer, or at a minimum strive towards answering, whether the large flow of domestic migration has a major effect on electoral behaviors.  

```{r}
library(tidyverse)
pres_2020 <- read_csv("pres_data/2020_US_County_Level_Presidential_Results.csv")
pres_2020 <- pres_2020 |> filter(state_name != "Alaska", state_name != "Connecticut", 
                                 state_name != "District of Columbia")
pres_2020 <- pres_2020 |> select(-votes_dem, -per_dem, -diff, -per_point_diff) |> 
  rename(votes_gop_2020 = votes_gop, per_gop_2020 = per_gop, total_votes_2020 = total_votes)

pres_2024 <- read_csv("pres_data/2024_US_County_Level_Presidential_Results.csv")
pres_2024 <- pres_2024 |> filter(state_name != "Alaska", state_name != "Connecticut", 
                                 state_name != "District of Columbia")
pres_2024 <- pres_2024 |> select(-votes_dem, -per_dem, -diff, -per_point_diff) |> 
  rename(votes_gop_2024 = votes_gop, per_gop_2024 = per_gop, total_votes_2024 = total_votes)

pop_2024 <- read_csv("pop_data/co-est2024-alldata.csv")
pop_2024 <- pop_2024 |> filter(STNAME != "Alaska", STNAME != "Connecticut", STNAME != "District of Columbia")
pop_2024 <- pop_2024 |> mutate(county_fips = paste0(STATE, COUNTY)) |> 
  mutate(dom_mig_total = (DOMESTICMIG2020 + DOMESTICMIG2021 + DOMESTICMIG2022 + DOMESTICMIG2023 + DOMESTICMIG2024)) |> 
  mutate (dom_mig_per = dom_mig_total / ESTIMATESBASE2020) |> mutate(population_2020 = ESTIMATESBASE2020)
```
```{r}
pop_2024_important <- pop_2024 |> select(county_fips, dom_mig_total, dom_mig_per, population_2020)

pop_2024_50_states <- pop_2024 |> filter(STNAME == CTYNAME) |> 
  arrange(desc(dom_mig_per)) |>
  select(STNAME, dom_mig_per, population_2020) 

percent_gop <- left_join(pres_2020, pres_2024) |> mutate(per_gop_change = per_gop_2024 - per_gop_2020)
percent_gop <- left_join(percent_gop, pop_2024_important, by = "county_fips")

focused_per_gop <- percent_gop |> select(county_fips, state_name, per_gop_change, dom_mig_per, population_2020)

percent_gop_50_states <- percent_gop |> group_by(state_name) |> 
  summarize(dom_mig_total = sum(dom_mig_total), population_2020 = sum(population_2020), gop_2020 = sum(votes_gop_2020),
            gop_2024 = sum(votes_gop_2024), total_2020 = sum(total_votes_2020), total_2024 = sum(total_votes_2024))

percent_gop_50_states <- percent_gop_50_states |> 
  mutate(domestic = (dom_mig_total / population_2020), gop_2024_per = (gop_2024 / total_2024), 
         gop_2020_per = (gop_2020 / total_2020)) |> 
  mutate(gop_change = (gop_2024_per - gop_2020_per))

percent_gop_50_states <- percent_gop_50_states |> mutate(dom_mig_per = dom_mig_total/population_2020) |> select(state_name, gop_change, dom_mig_per, population_2020) |>
  arrange(desc(dom_mig_per)) 


print(percent_gop_50_states)
#head(percent_gop)
#head(pop_2024)
```

```{r}
plot_basic <- ggplot(focused_per_gop, mapping = aes(x=dom_mig_per, y=per_gop_change)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE, linewidth = 2.0) + 
  labs(title = "The Effect of Domestic Migration on the 2024 Presidential Election", 
       x = "Net Proportional Domestic Migration per County, 2020-2024", 
       y = "Change in GOP Voteshare per County, 2020-2024", 
       caption = "Sources: 2024 PEP Census Data, 2020 and 2024 US House of Representatives Election Statistics")
plot_basic
```


```{r}
plot_50k <- focused_per_gop |> filter(population_2020 > 50000) |>
  ggplot(focused_per_gop, mapping = aes(x=dom_mig_per, y=per_gop_change)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE)
plot_50k
```
```{r}
plot_10k_50k <- focused_per_gop |> filter(population_2020 >= 10000 & population_2020 <= 50000) |>
  ggplot(focused_per_gop, mapping = aes(x=dom_mig_per, y=per_gop_change)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE)
plot_10k_50k
```

```{r}
plot_10k <- focused_per_gop |> filter(population_2020 < 10000) |>
  ggplot(focused_per_gop, mapping = aes(x=dom_mig_per, y=per_gop_change)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE)
plot_10k
```

